Neural Connectome of the Ctenophore Statocyst

### 2025-04-28 manuscript/Manuscript.qmd
fig2 and supple fig S2

Classical neural staining techniques did not provide clear images of neurons at the aboral organ. However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise_1973]. In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion. At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) <!--# figure.. -->. We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise_1973]. Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells <!--# figure of the distribution -->.

We reconstructed three aboral nerve-net (ANN) neurons, each with a syncytial morphology. Each ANN had multiple nuclei <!--# add numbers e.g. (2-5 nuclei) -->, with anastomosing membranes <!--# show example EM pictures -->. These neurons are distinct from the syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously identified by vEM in the body wall of *Mnemiopsis* [@Burkhardt2023]. Our results reveal a second type of syncytial neurons in ctenophores. <!--# describe the differences and also mention the classical SNNs that Sanja reconstructed -->

Furthermore, ANNs are also different from other neurons previously reported by EM, including mesogleal neurons and ciliated sensory cells that synapse on the SNN (types 1-4)[@Burkhardt2023]. Based on their distinct size and position, we classified the ANNs into two types. The first type is a single large neuron (ANN_Q1-4) spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei This neuron has X <!--# add number --> presynaptic sites. The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each spanning two quadrants and with two nuclei, and each containing X presynaptic structures. <!--#  -->

![**Figure 2. Organisation of the aboral synaptic nerve net**\
(A) 3D reconstruction of the aboral nerve net (ANN) Q1Q2Q3Q4 (Q1-4), which spans all four quadrants and contains multiple nuclei (blue). The spheres represent the positions of individual nuclei. The left panel shows a aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(B) 3D reconstruction of two ANN Q1Q2 (pink) and Q3Q4 (orange), each spanning two quadrants and containing multiple nuclei. The spheres indicate the positions of the nuclei. The left panel shows a aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(C) Graph showing the number of mitochondria per cell and the proportion of mitochondria involved in synapse formation for each cell type. Green represents mitochondria associated with synapses, while gray represents mitochondria not involved in synapse formation. ANN has a significantly higher number of mitochondria compared to other cell types, with most of them forming synapses. The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
(D) Localization of mitochondria within ANN Q1-4. Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.\
(E) Representative electron micrographs of presynaptic triad structures observed in the dataset and their schematic diagrams. The left diagram illustrates synaptic projections from the center ANN (ANN Q1-4) to the lateral ANN (ANN Q3Q4), while the right diagram shows synaptic projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN Q1-4). The abbreviations are as follows: mi (mitochondrion), er (endoplasmic reticulum), and sv (synaptic vesicles).\
(F) Plot showing the three-dimensional positions of synapses. Synapses from the center ANN to the lateral ANN are shown in magenta, while synapses from the lateral ANN to the center ANN are shown in cyan. Blue dots indicate the locations of autapses within ANN Q1-4.\
(G) Neural circuit map of ANN connectivity. The blue, orange, and pink circles represent individual neurons in ANN Q1-4, Q2Q3, and Q3Q4, respectively. Magenta arrows indicate synaptic connections from the center ANN to the lateral ANN, while cyan arrows indicate connections from the lateral ANN to the center ANN. Numbers denote the number of synapses. Blue arrows represent autapses in ANN Q1-4 along with their synapse counts.](figures/Figure2.png)

>>>>

Classical neural staining techniques did not provide clear images of neurons at the aboral organ. However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise_1973]. In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion. At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) (Supplementary Figure S2A). We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise_1973]. Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells (Supplementary Figure S2B).

We reconstructed three aboral nerve-net (ANN) neurons, each with a syncytial morphology (Figure 2A and B). Each ANN had multiple nuclei (ranging from two to five per neuron), with anastomosing membranes (Supplementary Figure S2C). These neurons are distinct from the syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously identified by vEM in the body wall of *Mnemiopsis* [@Burkhardt2023]. Our results reveal a second type of syncytial neurons in ctenophores (Supplementary Figure S2D).

Furthermore, ANNs are also different from other neurons previously reported by EM, including mesogleal neurons and ciliated sensory cells that synapse on the SNN (types 1-4)[@Burkhardt2023]. Based on their distinct size and position, we classified the ANNs into two types. The first type is a single large neuron (ANN_Q1-4) spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei. This neuron has 150 presynaptic sites. The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each spanning two quadrants and containing two nuclei. ANN_Q1Q2 had 119 presynaptic structures, whereas ANN_Q3Q4 had 68 presynaptic structures.

![**Figure 2. Organisation of the aboral synaptic nerve net**\
(A) 3D reconstruction of the aboral nerve net (ANN) Q1Q2Q3Q4 (Q1-4), which spans all four quadrants and contains multiple nuclei (blue). The spheres represent the positions of individual nuclei. The left panel shows a aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(B) 3D reconstruction of two ANN Q1Q2 (pink) and Q3Q4 (orange), each spanning two quadrants and containing multiple nuclei. The spheres indicate the positions of the nuclei. The left panel shows a aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(C) Graph showing the number of mitochondria per cell and the proportion of mitochondria involved in synapse formation for each cell type. Green represents mitochondria associated with synapses, while gray represents mitochondria not involved in synapse formation. ANN has a significantly higher number of mitochondria compared to other cell types, with most of them forming synapses. Except for lithocytes, each cell type has a sample size of at least three cells. The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
(D) Localization of mitochondria within ANN Q1-4. Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.\
(E) Representative electron micrographs of presynaptic triad structures observed in the dataset and their schematic diagrams. The left diagram illustrates synaptic projections from the center ANN (ANN Q1-4) to the lateral ANN (ANN Q3Q4), while the right diagram shows synaptic projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN Q1-4). The abbreviations are as follows: mi (mitochondrion), er (endoplasmic reticulum), and sv (synaptic vesicles).\
(F) Plot showing the three-dimensional positions of synapses. Synapses from the center ANN to the lateral ANN are shown in magenta, while synapses from the lateral ANN to the center ANN are shown in cyan. Blue dots indicate the locations of autapses within ANN Q1-4.](figures/Figure2.png)
